\documentclass{article}
\usepackage{geometry}
\usepackage{xcolor}
\usepackage{listings}
\lstset{%
  backgroundcolor=\color{gray!40},
  breaklines=true,
  basicstyle=\ttfamily,
  commentstyle=\ttfamily,
  language=fortran,
  showspaces=false,
  showstringspaces=false
}
\usepackage{enumerate}

\title{CLAW Compiler Documentation}
\author{Valentin Clement}
\date{\today\\\vspace{1em}v0.4}

\setlength\parindent{0pt}
\setlength{\parskip}{1em}

\begin{document}
\maketitle

CLAW Compiler is a source-to-source compiler interpreting the CLAW
directive language.

The CLAW Compiler is based on the OMNI Compiler. For more information
about the OMNI Compiler project, visit http://omni-compiler.org.

OMNI Compiler user manual can be found here: http://omni-compiler.org/manual.html

For more information about the CLAW directive language, visit
https://github.com/C2SM-RCM/claw-language-specification


\section{Compiler options}
\begin{lstlisting}
usage: clawfc_test <OPTIONS> <INPUTFILE> ...

CLAW Compiler options:

  -o <file>                  : place the output into <file>.
  -I <dir>                   : add the directory dir to the list of
                               directories to be searched for header
                               files.
  -J <dir>                   : specify where to put .mod and .xmod
                               files for compiled modules.
  -D <dir>                   : specify output directory for
                               transformed files
  -t=,--target=<target>      : specify the type of accelerator
                               directive language for code
                               generation.
  --target-list              : list the available type of
                               accelerator directive language
                               supported.
  -d=,--directive=<dir-lang> : specify the type of accelerator
                               directive language for code
                               generation.
  --directive-list           : list the available type of
                               accelerator directive language
                               supported.
  --config=<config_file>     : specify a different configuration for
                               the translator.
  --show-config              : list the current configuration
                               information. If used with --config,
                               list the information from the
                               specific configuration.
  -v,--verbose               : print processing status.
  --version                  : print version.
  -h,--help                  : print usage.
  --show-env                 : show environment variables.
  --no-dep                   : don't generate .mod or .xmod file for
                               dependencies.
  -f,--force                 : force the translation of files
                               without directives.
  --force-pure               : force compiler to exit when
                               transformation applied to PURE
                               subroutine/function.
  -r,--report                : generate the tranformation report.
  --debug                    : save intermediate files in
                               __omni_tmp__.
  --stop-pp                  : save intermediate files and stop
                               after preprocess.
  --stop-dependencies        : save intermediate files and stop
                               after dependencies resolution.
  --stop-frontend            : save intermediate files and stop
                               after frontend.
  --stop-translator          : save intermediate files and stop
                               after translator.

Decompiler options:
  -w <integer>               : Set the number of columns for the
                               output Fortran file (default: 80).
  -l                         : Add preprocessor line directives
                               in the output Fortran file.

Process Options

  --Wp[option] : Add preprocessor option.
  --Wf[option] : Add frontend option.
  --Wx[option] : Add Xcode translator option.
  --Wb[option] : Add backend option.
\end{lstlisting}

\subsection{Options for the output}
These options affect how the CLAW Compiler writes the output files.

\textbf{\texttt{-o}}\\
The output file can be specified with the \texttt{-o} option.
\begin{lstlisting}
$ clawfc -o transformed_code.f90 original_code.f90
\end{lstlisting}

\textbf{\texttt{-D}}\\
The option \texttt{-D} can be used to specify an output directory. If defined, all the files
will be written in the specified directory.

\begin{lstlisting}
$ clawfc -D src/ -o transformed_code.f90 original_code.f90
\end{lstlisting}

\subsection{Options for directory search}
These options affect how the CLAW Compiler searches for files specified by the
\texttt{"INCLUDE"} directive and where it searches for previously compiled modules.

\textbf{\texttt{-I}}\\
These affect interpretation of the \texttt{"INCLUDE"} directive (as well as of the \texttt{"\#include"} directive of the cpp preprocessor).
All the specified paths are used to search for header files.
\begin{lstlisting}
$ clawfc -I mpi/ -o transformed_code.f90 original_code.f90
\end{lstlisting}

\textbf{\texttt{-J}}\\
All specified paths are used to search for \texttt{.mod/.xmod} file when previously compiled modules are required by a \texttt{"USE"} statement.
\begin{lstlisting}
$ clawfc -J xmod/ -o transformed_code.f90 original_code.f90
\end{lstlisting}

\subsection{Options to affect the workflow}
These options affect the workflow of the compiler.

\textbf{\texttt{--no-dep}}\\
Do not generate module file (\texttt{.xmod}) for the dependencies. By default, module file for the dependencies are generated automatically.
\begin{lstlisting}
$ clawfc --no-dep -o transformed_code.f90 original_code.f90
\end{lstlisting}

\textbf{\texttt{-f,--force}}\\
Force the compilation of files that do not have CLAW directives. By default, files without CLAW directives are ignored and just copied to the destination file.
\begin{lstlisting}
$ clawfc --force -o transformed_code.f90 original_code.f90
\end{lstlisting}

\textbf{\texttt{--stop-pp}}\\
Stop the compiler after the preprocessing.
\begin{lstlisting}
$ clawfc --stop-pp -o transformed_code.f90 original_code.f90
\end{lstlisting}

\textbf{\texttt{--stop-dependencies}}\\
Stop the compiler after the dependencies resolution. Module files are generated but nothing is done with the input file.
\begin{lstlisting}
$ clawfc --stop-dependencies -o transformed_code.f90 original_code.f90
\end{lstlisting}

\textbf{\texttt{--stop-frontend}}\\
Stop the compiler after the front-end. Input XcodeML file is generated.
\begin{lstlisting}
$ clawfc --stop-frontend -o transformed_code.f90 original_code.f90
\end{lstlisting}

\textbf{\texttt{--stop-translator}}\\
Stop the compiler after the translation. Output XcodeML file is generated from the input XcodeML file with applied translation.
\begin{lstlisting}
$ clawfc --stop-translator -o transformed_code.f90 original_code.f90
\end{lstlisting}

\subsection{Options for the accelerator language directive}
These options affect the generation of accelerator language directive.

\textbf{\texttt{-d=,--directive=<dir-lang>}}\\
Allow to specify the type of accelerator directive language for code generation.

\textbf{\texttt{--directive-list}}\\
Print the list of available type of accelerator directive language supported.

\subsection{Options for the code generation target}
These options affect the way the code generation is performed depending on the target.

\textbf{\texttt{-t=,--target=<target>}}\\
Allow to specify the type of accelerator directive language for code generation.

\textbf{\texttt{--target-list}}\\
Print the list of available type of accelerator directive language supported.

\subsection{Options for the global configuration}
Default target and directive language as well as transformation order are
specified by a configuration file.
A default file is generated an installed with the CLAW Compiler.
An option allows to use a different configuration while using the compiler.

\begin{lstlisting}
  <!-- CLAW default configuration. This file should not be edited ! -->
  <claw version="1.0">
    <!-- Transformation parameters -->
    <global>
      <!-- Default general values -->
      <parameter key="default_target" value="gpu"/>
      <parameter key="default_directive" value="openacc"/>

      <!-- OpenACC default information -->
      <!-- Define the default values to be used for the clauses. If value set to 0, compiler default value used. -->
      <parameter key="acc_vector_length" value="128"/> <!-- vector_length(128) -->
      <parameter key="acc_num_gangs" value="8"/>       <!-- num_gangs(8) -->
      <parameter key="acc_num_workers" value="8"/>     <!-- num_workers(8) -->
      <!-- Specify the order of generation of clauses when nested loops are generated -->
      <!-- possible values are vector_gang, gang_vector, vector, gang, none -->
      <parameter key="openacc_execution_mode" value="vector_gang"/>
    </global>
    <!-- Transformation groups -->
    <groups>
      <!-- Low-level transformations -->
      <group name="remove" />
      <group name="directive-primitive" />
      <group name="array-transform" />
      <group name="loop-extract" />
      <group name="loop-hoist" />
      <group name="loop-fusion" />
      <group name="loop-interchange" />
      <group name="on-the-fly" />
      <group name="kcache" />
      <group name="if-extract" />
      <!-- High-level transformations -->
      <group name="sca" />
      <group name="sca-forward" />
      <!-- internal applied at the end -->
      <group name="openacc-continuation" />
    </groups>
  </claw>
\end{lstlisting}

\textbf{\texttt{--config=<config-file>}}\\
Allow to specify a different configuration for the translator.

\textbf{\texttt{--show-config}}\\
Print the the current configuration information. If used with \texttt{--config}, list the information from the specific configuration.

\subsection{Decompiler options}
\textbf{\texttt{-w <integer>}}\\
Set the number of columns for the output Fortran file (default: 80).

\textbf{\texttt{-l}}\\
Add preprocessor line directives in the output Fortran file.

\subsection{Options for debugging}
These options affect the output at the command line and give information about the process.

\textbf{\texttt{-v,--verbose}}\\
Print the inner commands run during the compilation process.

\textbf{\texttt{--version}}\\
Print the version of the compiler.

\textbf{\texttt{-h,--help}}\\
Print help about the compiler.

\textbf{\texttt{--show-env}}\\
Show all the environment variables used in the compiler workflow.

\textbf{\texttt{--debug}}\\
Print process information and store all the intermediate files in \_\_omni\_tmp\_\_ directory.

\subsection{Options for subprocesses}
These options affect the different process part of the CLAW Compiler.

\textbf{\texttt{--Wp}}\\
Pass options to the underlying preprocessor.
\begin{lstlisting}
$ clawfc --Wp-D_OPENACC -o transformed_code.f90 original_code.f90
\end{lstlisting}

\textbf{\texttt{--Wf}}\\
Pass options to the underlying front-end.
\begin{lstlisting}
$ clawfc --Wf-no-module-cache -o transformed_code.f90 original_code.f90
\end{lstlisting}

\textbf{\texttt{--Wd}}\\
Pass options to the underlying translator.
\begin{lstlisting}
$ clawfc --Wd-d -o transformed_code.f90 original_code.f90
\end{lstlisting}

\textbf{\texttt{--Wb}}\\
Pass options to the underlying preprocessor.
\begin{lstlisting}
$ clawfc --Wb-w80 -o transformed_code.f90 original_code.f90
\end{lstlisting}

\end{document}
