\chapter{Transformation}
\label{chapter:transformation}
A transformation is the the basic representation of a manipulation of the
\gls{ast} triggered by a directive or for the whole translation unit. Each
translation unit transformation or directive transformation is implemented in
its own transformation class. If the directive can be used as a block directive
(with a \lstinline|!$end <directive>| directive), the transformation class must
inherits from the \lstinline!ClawBlockTransformation! super class. Otherwise,
it can inherits from the basic \lstinline!ClawTransformation! super class.

\section{Type of transformation}
\label{section:trans_type}
Transformations are divided into two distinct groups. The independent and the
dependent transformations. The first one, as its name implies, is performed
independently of any other transformations. Translation unit transformations
are always independent.
The dependent transformation, in the
other hand, is applied only when it can be combined with another dependent
transformation of the same kind in its group. Most of the transformations are
independent. The best example for the dependent transformation is the loop
fusion transformation. As shown in Listing~\ref{lst:trans_dep}, there are two
CLAW directives (line 4 and 9). These directives will both trigger a dependent
loop fusion transformations. Alone, those transformations have no effect
on the \gls{ast}. The translation engine will analyze if the
first one can be transformed with the second one. If so, the transformations
will be grouped together and the fusion will be done. Otherwise, the
transformations are just ignored as they have to depend on at least one other
transformation. Checks that match dependent transformation together are
specific to the kind of transformation.

\lstinputlisting
  [
    label=lst:trans_dep,
    caption=Loop fusion example,
    language=Fortran
  ]{code/trans_dependent.f90}

\section{Trigger transformation}
\label{section:trans_trigger}
Transformations can be triggered by a directive or they can be applied for each
translation unit independently from the directives present in it or not.
This information is specified in the configuration file of the transformation
set (see Section~\ref{section:conf_trans_set}). Directive triggered
transformations must implement the constructor passing the
\lstinline|ClawPragma| parameter.


\section{Transformation application order}
The first step in the translation of a translation unit (an \xcodeml \gls{ir}),
is the detection of all the directives. Each directive will trigger the
creation of an instance of the transformation class it belongs to, if any.
For example, a loop-fusion directive will trigger the creation of a
\lstinline|LoopFusion| instance. On this instance, the \lstinline|analyze()|
method is called in order to determine if the transformation can take place.
If the analysis step is successful, the transformation is added to its
transformation group. Once all the transformation instances have been analyzed
and categorized by groups, the actual code transformation can take place on
the \xcodeml \gls{ir}. All transformations in a group are applied one after
another in a \textit{fifo} order. The order in which groups are processed is
determined by the configuration file.

\lstinputlisting
  [
    label=lst:config,
    caption=CLAW default configuration,
    language=xml
  ]{code/claw-default.xml}

As shown in Figure~\ref{lst:config}, the transformation order is specified the
configuration file under the XML format.
Each transformation group name refers to its name in the transformation set it
belongs to. Global configuration and transformation set file are described in
details in Chapter~\ref{chapter:configuration}.

\section{Add a new transformation}
\label{section:new_trans}
A transformation in the \clawfc is represented as a class that inherits from
the \lstinline|ClawTranformation| or \lstinline|ClawBlockTransformation|
super class. In order to add a new transformation into the the CLAW Compiler,
the following steps must be done.

\begin{enumerate}
\item Create a new class that inherits from one of the transformation
      super class.
\item \textit{(Optional)} Define the directive that will trigger the
      transformation and add it to the CLAW directive language parser if the
      transformation is triggered by a directive.
\item \textit{(Optional)} Detect and categorize the new transformation if the
      transformation is triggered by a directive in the translator.
\item Add the transformation to a transformation set configuration and to
      the global configuration.
\end{enumerate}

As an example, the next 4 subsection describe those steps with more details.

\subsection{New transformation class}
The transformation created at this step will be a simple independent
transformation triggered by a directive. It will then inherits from
\lstinline|ClawTransformation| super class.

\lstinputlisting
  [
    label=lst:myfirsttransformation.java,
    caption=MyFirstTransformation.java,
    language=java
  ]{code/MyFirstTransformation.java}

The transformation class shown in Listing~\ref{lst:myfirsttransformation.java}
is really simple. It just inherits from \lstinline|ClawTransformation| on line
16. Therefore, it has to implements the \lstinline|analyze()|,
\lstinline|transform()| and \lstinline|canBeTransformedWith()| methods as they
are abstract in the super class.

The \lstinline|analyze()| method does not perform any check and just return
\lstinline|true| to tell the translation engine that the transformation can
be applied.

The \lstinline|transform()| method is pretty simple. It delete the pragma that
triggered the transformation from the \gls{ast}. It will then not be in the
resulting transformed code.

\subsection{Directive and CLAW directive language parser}
If the new transformation class is triggered by a directive, it needs to
be set up in the directive language parser in order to be instantiated.
Listing~\ref{lst:mydirective} is the directive that is used in
the current example.

\begin{lstlisting}[label=lst:mydirective, caption=Example directive,
  language=fortran]
!$claw mydirective
\end{lstlisting}

To parse the directive, two files need to be modified. First, the new directive
needs to be added to the \lstinline|ClawDirective| enumeration. It is added as
the last element of the enumeration in the Listing~\ref{lst:clawdirective}.

\begin{lstlisting}[label=lst:clawdirective, caption=ClawDirective.java,
  language=java]
package cx2t.wani.language;

public enum ClawDirective {
  ARRAY_TRANSFORM,
  ARRAY_TO_CALL,
  DEFINE,
  IGNORE,
  KCACHE,
  LOOP_FUSION,
  LOOP_INTERCHANGE,
  LOOP_HOIST,
  LOOP_EXTRACT,
  PRIMITIVE,
  SCA,
  REMOVE,
  VERBATIM,
  MYDIRECTIVE
}
\end{lstlisting}

Then, the CLAW directive language parser has to be modified to understand
this new directive and return the correct value from the enumeration.

\begin{lstlisting}[label=lst:clawdirectiveantlr, caption=Claw.g4,
  language=java]
grammar Claw;

@header
{
import cx2t.wani.ClawConstant;
import cx2t.wani.language.*;
import cx2t.tatsu.common.*;
}

/*----------------------------------------------------------------------------
 * PARSER RULES
 *----------------------------------------------------------------------------*/

// Entry point of the parsing
analyze returns [ClawPragma cp]
  @init{ $cp = new ClawPragma(); }:
  CLAW directive[$cp] EOF
;

directive[ClawPragma cp]:
  // The new directive
  MYDIRECTIVE { $cp.setDirective(ClawDirective.MYDIRETIVE); }
;

/*----------------------------------------------------------------------------
 * LEXER RULES
 *----------------------------------------------------------------------------*/

// Start point
CLAW         : 'claw';
MYDIRECTIVE  : 'mydirective';
\end{lstlisting}

In Listing~\ref{lst:clawdirectiveantlr}, there is a minimal version of the
ANTLR grammar file of the CLAW directive language parser. The token
\lstinline|mydirective| is added in the lexer rules (line 31) and is then
used in the parser rules (line 22). ANTLR grammar file accepts Java code
to be inserted during the parsing. On line 22, the enumeration value for
the new directive is set in the analyzed pragma object.

\subsection{Detection and categorization}
The new directive has its transformation class and can be analyzed by the
CLAW directive language parser.

Now, it needs to be detected in order to create the transformation instance.
This happens in the \lstinline|ClawTranslator| class. The
\lstinline|generateTransformation()| method detects all the pragmas in the
translation unit and categorize them.

As shown in Listing~\ref{lst:categorization}, a new case is added to the switch
statement that categorize the transformation. Here, a new instance of
\lstinline|MyFirstTransformation| transformation is created and added to the
correct group as long as its analysis step succeed.

\begin{lstlisting}[label=lst:categorization, caption=ClawTranslator.java,
  language=java]
switch(analyzedPragma.getDirective()) {
  case MYDIRECTIVE:
    addTransformation(xcodeml, new MyFirstTransformation(analyzedPragma));
    break;
}
\end{lstlisting}


\subsection{Enable the new transformation}
Finally, the new transformation needs to be enabled by adding it to the
configuration file.

Listing~\ref{lst:new_trans_set} shows the transformation set configuration
including the new transformation class.


\begin{lstlisting}[label=lst:new_trans_set, caption=claw-dummy-set.xml, language=xml]
<transformations>
  <transformation name="delete-pragma" type="independent" trigger="directive"
    class="cx2t.wani.transformation.MyFirstTransformation"/>
</transformations>
\end{lstlisting}

Listing~\ref{lst:new_trans_conf} uses the transformation set from
Listing~\ref{lst:new_trans_set} and enables its only transformation.

\begin{lstlisting}[label=lst:new_trans_conf, caption=claw-dummy.xml, language=xml]
<claw version="1.0">
  <global type="extension"/>
  <sets>
    <set name="claw-dummy"/>
  </sets>
  <groups>
    <group name="delete-pragma"/>
  </groups>
</claw>
\end{lstlisting}

\subsection{Test the new transformation}
The last step is to compile and install the new version of the \clawfcomp
containing the new directive and transformation class. Once it is done, it
can be directly tested with a simple Fortran program.

Listing~\ref{lst:mydirectiveoriginal} is the original code. Once it is
transformed by the CLAW Compiler, it will produce the code
in Listing~\ref{lst:mydirectivetransformed}. The command to call
the \clawfcomp is shown in Listing~\ref{lst:clawfcexample}.

\lstinputlisting
  [
    label=lst:mydirectiveoriginal,
    caption=original.f90,
    language=fortran
  ]{code/mydirective_orig.f90}

\lstinputlisting
  [
    label=lst:mydirectivetransformed,
    caption=transformed.f90,
    language=fortran
  ]{code/mydirective_trans.f90}

\begin{lstlisting}[label=lst:clawfcexample, caption=Call the compiler, language=bash]
$ clawfc -o transformed.f90 original.f90
\end{lstlisting}
